---
title: 'Omic Data Analysis PEC1'
author: 'Laura Sudupe Medinilla'
subtitle: '`r params$subtitulo`'
date: '`r format(Sys.Date(),"%e de %B, %Y")`' 
# date: \today  (solo para pdf)
abstract:
output:
  pdf_document:
    keep_tex: yes
    toc: yes
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    theme: united
    higlight: tango
header-includes: \usepackage[spanish]{babel}
params:
  file1: genotype.csv
  file2: flowering_time.csv
  folder.data: ./floweringTime
  b: 40
  p.train: 0.66667
  subtitulo: Analysis of microarray data
  valor.seed: 12345
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NULL)
options(width=90)
```

# Objetives
# Methods and Materials
## Software

```{r libraries, warning=FALSE, message=FALSE, echo=TRUE}

```






## Data
## Analysis

### Identification of the groups and the samples groups
### Exploration and quality control of the raw data
### Normalization
### Quality control of the processed data (optional)
### No-especific filtering (optional)
### Identification of the differencialy expressed genes
### Results annotations
### Diferent comparation comparations
### Biologic signification analysis ("Gene Enrichment Analysis")




# Results
# Discusion
# Apendice
```{r packages, message=FALSE, echo=FALSE, warning=FALSE}
libraries <- c("e1071", "gmodels",  "ROCR", "caret")
check.libraries <- is.element(libraries, installed.packages()[, 1])==FALSE
libraries.to.install <- libraries[check.libraries]
if (length(libraries.to.install!=0)) {
  install.packages(libraries.to.install)
}
library("e1071")
library("gmodels")
library("ROCR")
library("caret")
```


\pagebreak




