---
title: "Datos ómicos unidad 1"
author: "Laura Sudupe Medinilla"
date: "27/2/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

```



# EXERCISE 1. expressionSet vignette

- Get the ExpressionSet vignette
- Create of an expressionSet from scratch
- Practice access and modification of slot contents, expeacially using exprs() 
and pData()


The definition of the **ExpressionSet** class along with many methods for 
manipulating **ExpressionSet** objets are defined in the *Biobase* package.
```{r}
#install.packages("BiocManager")
#BiocManager::install("Biobase")
library("Biobase")
```


The data from many high-throughput genomic experiments, such as microarray 
experiments, usually consist of several conceptually distinct parts: assay data, 
phenotypic meta-data, feature annotations and meta-data, and a description of 
the experiment. We´ll construct each of these components, and then assemble them 
into an **ExpressionSet**.

## Assay data

Matrix of "expression" values from the experiment. The values usually derived 
from microarrays of one sort or another, perphaps after initial processing by 
manufactuer software or Bioconductor packages.

The matrix has F rows and S columns, where *F* is the number of features on the 
chip and *S* is the number of samples.

```{r}
dataDirectory <- system.file("extdata", package="Biobase")
exprsFile <- file.path(dataDirectory, "exprsData.txt")

#These first two lines create a file path pointing to where the assay data is 
#stored

exprs <- as.matrix(read.table(exprsFile, header=TRUE, sep="\t",
                              row.names=1,
                              as.is=TRUE))
```

Lets verify that the data have read matches your expectations. At minimun, check
the class and dimensions of **geneData** and take a peak at the first several 
rows

```{r}
class(exprs)
```
```{r}
dim(exprs)
```
```{r}
colnames(exprs)
```
```{r}
head(exprs[,1:5])
```
We create a minimal *ExpressionSet* object using the **ExpressionSet** 
constructor:

```{r}
minimalSet <- ExpressionSet(assayData = exprs)
```










# EXERCISE 2. GEOquery package



