---
title: "Datos ómicos unidad 1"
author: "Laura Sudupe Medinilla"
date: "27/2/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# EXERCISE 1. expressionSet vignette

- Get the ExpressionSet vignette
- Create of an expressionSet from scratch
- Practice access and modification of slot contents, expeacially using exprs() 
and pData()


The definition of the **ExpressionSet** class along with many methods for 
manipulating **ExpressionSet** objets are defined in the *Biobase* package.
```{r}
#install.packages("BiocManager")
BiocManager::install("Biobase", version = "3.8")
```


The data from many high-throughput genomic experiments, such as microarray 
experiments, usually consist of several conceptually distinct parts: assay data, 
phenotypic meta-data, feature annotations and meta-data, and a description of 
the experiment. We´ll construct each of these components, and then assemble them 
into an **ExpressionSet**.
```{r}

```












# EXERCISE 2. GEOquery package



