---
title: "Datos ómicos unidad 1"
author: "Laura Sudupe Medinilla"
date: "27/2/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

```



# EXERCISE 1. expressionSet vignette

- Get the ExpressionSet vignette
- Create of an expressionSet from scratch
- Practice access and modification of slot contents, expeacially using exprs() 
and pData()


The definition of the **ExpressionSet** class along with many methods for 
manipulating **ExpressionSet** objets are defined in the *Biobase* package.
```{r}
#install.packages("BiocManager")
#BiocManager::install("Biobase")
library("Biobase")
```


The data from many high-throughput genomic experiments, such as microarray 
experiments, usually consist of several conceptually distinct parts: assay data, 
phenotypic meta-data, feature annotations and meta-data, and a description of 
the experiment. We´ll construct each of these components, and then assemble them 
into an **ExpressionSet**.

## Assay data

Matrix of "expression" values from the experiment. The values usually derived 
from microarrays of one sort or another, perphaps after initial processing by 
manufactuer software or Bioconductor packages.

The matrix has F rows and S columns, where *F* is the number of features on the 
chip and *S* is the number of samples.

```{r}
dataDirectory <- system.file("extdata", package="Biobase")
exprsFile <- file.path(dataDirectory, "exprsData.txt")

#These first two lines create a file path pointing to where the assay data is 
#stored

exprs <- as.matrix(read.table(exprsFile, header=TRUE, sep="\t",
                              row.names=1,
                              as.is=TRUE))
```

Lets verify that the data have read matches your expectations. At minimun, check
the class and dimensions of **geneData** and take a peak at the first several 
rows

```{r}
class(exprs)
```
```{r}
dim(exprs)
```
```{r}
colnames(exprs)
```
```{r}
head(exprs[,1:5])
```
We create a minimal *ExpressionSet* object using the **ExpressionSet** 
constructor:

```{r}
minimalSet <- ExpressionSet(assayData = exprs)
```


## Phenotypic data

Phenotypic data summarizes information about samples (sex, age, and treatment 
status;refererd to as "covariates"). The information describing the samples can 
be represented as a table with *S* rows and *V* columns, where *v* is the number
of covariates.

```{r}
pDataFile <- file.path(dataDirectory, "pData.txt")
pData <- read.table(pDataFile, 
                    row.names=1,
                    header=TRUE,
                    sep="\t")
                    
dim(pData)
```
```{r}
rownames(pData)
summary(pData)
```
Note thar the number of rows of phenotypic data amtch the number of columns of 
expression data, and indeed that the row and columns names are identically ordered.
```{r}
all(rownames(pData)==colnames(exprs))
```
Phenotypic data can take a lot of different forms. Some covariates might be
represented as numeric values. Others (gender, tissue type, cancer status) might
better be represented as **factor** objets. It is especially important that the
phenotypic data are encoded correctly.

```{r}
#We can deermine the classes of each column with sapply
sapply(pData, class)
```
What is the sex and Case/Control status of the 15th and 20th samples? And for 
the sample(s) with score greater than 0.8.
```{r}
pData[c(15,20), c("gender", "type")]
pData[pData$score>0.8,]
```
When the column names doesn´t provide enough information about the covariate
we can create a dataframe containing such meta-data, or read the information from
a file using read.table
```{r}
metadata <- data.frame(labelDescription=
                         c("Patient gender",
                           "Case/control status",
                           "Tumor progress on XYZ scale"),
                       row.names = c("gender", "type", "score"))
```

Bioconductor´s **Biobase** package provides a class called *AnnotatedDataFrame**
that conveniently stores and manipulates the phenotypic data and its metadata in 
a coordinated fashion
```{r}
phenoData <- new("AnnotatedDataFrame", 
                 data=pData, varMetadata=metadata)
```

This *AnnotatedDataFrame* objects can be subset much like a *dataframe*
```{r}
head(pData(phenoData))
phenoData[c("A", "Z"), "gender"]
```

## Annotations and feature data

Meta-data on features is as important as meta-data on samples, the idea is to 
construct specialized meta-data packages for each type of chip or instrument.
There are a lot of them availables in Bioconductor, also, there are other meta
data packages provides by other initiatives such as GO and KEEG.
```{r}
annotation <- "hgu95av2"
```



## Experiment description

Basic description about the experiment (the investigator or lab, title) can be
recorded by creating a *MIAME* object.
```{r}
experimentData <- new("MIAME",
                      name="Pierre Fermat",
                      lab="Francis Galton Lab",
                      contact="pfermat@lab.not.exist",
                      title="Smoking-Cancer Experiment",
                      abstract="An example ExpressionSet",
                      url="www.lab.not.exist",
                      other=list(
                      notes="Created from text files"
                      ))
```


## Assembling an *ExpressionSet*

An *ExpressionSet* object is created by assembling its component parts and 
calling the **ExpressionSet** constructor
```{r}
exampleSet <- ExpressionSet(assayData=exprs,
                            phenoData=phenoData,
                            experimentData = experimentData,
                            annotation = "hgu95av2")
```


# EXERCISE 2. GEOquery package

- Install and load the GEOquery package
  - Browse te GEPquery vignette
- Create an ExpressionSet by downloading a dataset from GEO
- Investigate the structure of the object
- Find out how you can access and change
    - The expression values or
    - the covariates in the phenoData

```{r}
#download Geoquery
#BiocManager::install("GEOquery")
#library("GEOquery")
#browseVignettes("GEOquery")
```

Get the data from GEO
```{r}
gse161435 <- getGEO("GSE161435", GSEMatrix = TRUE)
```
```{r}
show(gse161435)
```
## Create a ExpressionSet

GEO Series are collections of related experiments. The getGEO() function can 
handle this format and will parse very large GSEs returning a lis of ExpressionSets
```{r}
show(pData(phenoData(gse161435[[1]]))[1:6, c(1,6,8)])
```

First we need to make sure that all of the GSM() are from the same platform

```{r}
gse <- getGEO("GSE161435", GSEMatrix = FALSE)
head(Meta(gse))
```
```{r}
# names of all the GSM objets contained in the GSE
names(GSMList(gse))

#get the first GSM object on the list
GSMList(gse)[[1]]
```
```{r}
# and the names of the GLPs represented
names(GPLList(gse))
```


```{r}
gsmplatforms <- lapply(GSMList(gse), function(x) {Meta(x)$platform_id})
head(gsmplatforms)
```
I only have one GLP, GLP21825, the plataforms we have determined by looking at
the GPLList for gse. We can filter the original GSMList to include only those
GSMs with the GLPxxx platform and use this list for further processing.

```{r}
gsmlist = Filter(function(gsm) {Meta(gsm)$platform_id=='GPL21825'},GSMList(gse))
length(gsmlist)
```
Now we would like to know what column represents the data that we would like to 
extract. GEO uses a convention that the column that contains the single measurement
for each array is called the VALUE column, which we could use if we don´t know 
what other column is most relevant
```{r}
Table(gsmlist[[1]])[1:6,]

#and get the column descriptions
Columns(gsmlist[[1]])[1:6,]
```
We will indeed use the VALUE() column
```{r}
# get the probeset ordering
probesets <- Table(GPLList(gse)[[1]])$ID

# make the data matrix from the VALUE columns from each GSM
# being careful to match the order of the probesets in the platform
# with those in the GSMs
data.matrix <- do.call('cbind',lapply(gsmlist,function(x) 
                                      {tab <- Table(x)
                                       mymatch <- match(probesets,tab$ID_REF)
                                       return(tab$VALUE[mymatch])
                                     }))
data.matrix <- apply(data.matrix,2,function(x) {as.numeric(as.character(x))})
data.matrix <- log2(data.matrix)
data.matrix[1:5,]
```
Note that we do a match() to make sure that the values and the platform 
information are in the same order.

Finally, to make the ExpressionSet()
```{r}
require(Biobase)

# go through the necessary steps to make a compliant ExpressionSet
rownames(data.matrix) <- probesets
colnames(data.matrix) <- names(gsmlist)
pdata <- data.frame(samples=names(gsmlist))
rownames(pdata) <- names(gsmlist)
pheno <- as(pdata,"AnnotatedDataFrame")
eset2 <- new('ExpressionSet',exprs=data.matrix,phenoData=pheno)
eset2
```





